---
title: "NHS Severe Patient Harm Forecasting"
author: "Emmanuel Oparaku & Michael Bakalov"
date: "July 2026"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{css, echo=FALSE}
caption {
  color: #666666;
  font-size: 0.9em;
  font-style: italic;
}

p.caption {
  color: #666666;
  font-size: 0.9em;
  font-style: italic;
}
```

## Introduction

This report describes a machine learning pipeline for forecasting daily
estimated avoidable deaths across the Bristol, North Somerset, and South
Gloucestershire (BNSSG) NHS system over rolling 10-day windows. The
pipeline uses Random Forest regression with direct multi-horizon
forecasting — one model trained per horizon from 1 to 10 days ahead.
This approach is preferred over recursive forecasting because each model
is trained directly against its target lead time, avoiding compounding
prediction errors across horizons.

The development dataset covers March 2023 to September 2025 (\~16.7M
rows, long format). The assessment period runs from 1 October 2025 to 17
February 2026, yielding 131 sliding 10-day forecast periods.

------------------------------------------------------------------------

## Feature Engineering

Feature selection prioritised clinical distinctiveness over raw
correlation rank. Rather than taking the top-N predictors by Spearman
correlation, we curated 10 predictors across four clinical domains —
community NCtR capacity, acute-to-community transfer backlog, acute
hospital pressure, and emergency demand — then collapsed correlated
predictors within each domain into derived aggregates. This avoids
feeding redundant signals to the model while preserving each domain's
contribution.

A key design constraint is the 3-day reporting lag on the outcome: only
lags of 3 days or more are observable at prediction time. This
constraint explains a non-monotonic pattern in CV MSE by horizon — MSE
peaks at horizon 3 (the lag boundary) and then decreases slightly as the
model shifts reliance from outcome lags toward clinical predictor
features. Feature selection itself used maximum absolute Spearman
correlation against the outcome at horizons h=3:10, computed on pre-2025
data only to prevent leakage.

------------------------------------------------------------------------

## Model Selection

Five models were evaluated through 8-fold expanding-window
cross-validation with 10-day validation windows and a 10-day gap between
training and validation:

```{r model-comparison, echo=FALSE}
results <- data.frame(
  Model = c("Random Forest", "XGBoost", "Linear Regression",
            "Rolling Mean (3-day)", "Naive (same weekday)"),
  `MSE (1-5)` = c(0.083, 0.095, 0.114, 0.111, 0.153),
  `SD (1-5)`  = c(0.066, 0.072, 0.059, 0.051, 0.040),
  `MSE (6-10)` = c(0.079, 0.089, 0.109, 0.180, 0.051),
  `SD (6-10)`  = c(0.034, 0.039, 0.054, 0.131, 0.054),
  check.names = FALSE
)
knitr::kable(results,
  caption = "Table 1: Cross-validated MSE by forecast horizon.
             Random Forest outperforms all baselines across both
             horizon groups. Lower is better.",
  digits = 3)
```

Random Forest won both horizon groups cleanly, beating the next best
model (XGBoost) by 13% on days 1–5 and 11% on days 6–10. The gap over
the naive same-weekday baseline is substantial — 46% on days 1–5 —
confirming that the clinical predictor features add genuine predictive
signal beyond simple calendar patterns. XGBoost was competitive but less
consistent across folds, with higher standard deviation. Random Forest
was therefore selected for submission and refit on all 927 development
rows.

------------------------------------------------------------------------

## Residual Analysis and Bias Correction

Out-of-fold residuals revealed two systematic patterns. First, a weekly
cycle: Saturday predictions were consistently lower than actuals (+0.093
correction needed), while Thursday predictions were closest to actuals
(+0.026 correction needed). Second, a seasonal bias: January was
persistently underpredicted, driven by the post-holiday surge period in
fold 5 (January 2025).

```{r residual-dow-plot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure 1: Mean out-of-fold residual by day of week for Random Forest and XGBoost. All days show systematic underprediction (negative residuals). Saturday is worst for Random Forest (-0.093), motivating the day-of-week bias correction."}
library(dplyr)
library(tidyr)
library(ggplot2)

resid_df <- readRDS("results/resid_df.rds")

resid_df %>%
  group_by(dow) %>%
  summarise(
    mean_rf  = mean(rf_resid,  na.rm = TRUE),
    mean_xgb = mean(xgb_resid, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  pivot_longer(c(mean_rf, mean_xgb),
               names_to  = "model",
               values_to = "mean_resid") %>%
  mutate(model = recode(model,
                        mean_rf  = "Random Forest",
                        mean_xgb = "XGBoost")) %>%
  ggplot(aes(x = dow, y = mean_resid, fill = model)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("Random Forest" = "#E07A5F",
                               "XGBoost"       = "#3AAFA9")) +
  labs(x = NULL, y = "Mean residual", fill = NULL) +
  theme_minimal()
```

These patterns motivated two additive post-hoc corrections, both
estimated from out-of-fold residuals only. The day-of-week correction is
applied at the **target date** for each horizon — the date the
prediction actually lands on — not the origin date. This is the
principled choice because the weekly cycle is a property of the day
being predicted, not the day forecasting starts. The monthly correction
is applied at the origin date, uniformly across all 10 horizons,
capturing residual seasonal signal not explained by the calendar
features.

**However, on the actual assessment data, the correction worsened MSE:**

```{r bias-correction, echo=FALSE}
bias_results <- data.frame(
  Horizon = c("Days 1-5", "Days 6-10"),
  `MSE (Raw)` = c(0.098, 0.110),
  `MSE (Corrected)` = c(0.103, 0.126),
  `Improvement` = c(-0.005, -0.016),
  `% Improvement` = c("-4.7%", "-14.5%"),
  check.names = FALSE
)
knitr::kable(bias_results,
  caption = "Table 2: Assessment period MSE — bias correction worsens
             performance across both horizon groups. Negative improvement
             indicates the correction increases error.",
  digits = 3)
```

The correction overcorrects on the assessment window probably because it
was estimated from predominantly non-winter development data (March 2023
– September 2025), yet the assessment period is entirely winter (October
2025 – February 2026). The model already predicts higher values in
winter through its clinical features; the correction then pushes
predictions further up, overshooting. This is consistent with the known
limitation of post-hoc corrections estimated from seasonally unbalanced
training data.

We therefore submit **uncorrected predictions as primary**
(`submission/pred_matrix.csv`) and bias-corrected predictions as
supplementary (`submission/pred_matrix_corrected.csv`). Both files are
generated automatically by `05_forecast.R`, with the MSE comparison
running before the save step so the rationale is documented in the
console output on every run.

------------------------------------------------------------------------

## Known Limitations

The most significant limitation is fold 5 (January 2025 post-holiday
surge), which dominates CV variance and is largely irreducible. The
model has seen only one January surge in training, and even with
explicit holiday flags, Random Forest's averaging mechanism compresses
extreme predictions toward the centre of the distribution —
underpredicting on high-mortality days and overpredicting slightly on
low-mortality days. This mean reversion is visible in the actual vs
predicted scatter at horizon 1 and explains why the January bias
correction, estimated from fold 5 alone, carries high uncertainty.

A secondary data quality issue: origin date 2025-10-03 is absent from
the assessment dataset, so period 4 (predicting October 4–13) produces
NA predictions, excluded from MSE calculations via `na.rm = TRUE`.
